---
title: "Market Basket Analysis"
author: "Ursula Podosenin"
date: "2024-07-03"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Imagine 10000 receipts sitting on your table. Each receipt represents a transaction with items that were purchased. The receipt is a representation of stuff that went into a customer’s basket – and therefore ‘Market Basket Analysis’.

That is exactly what the Groceries Data Set contains: a collection of receipts with each line representing 1 receipt and the items purchased. Each line is called a transaction and each column in a row represents an item.

The data set (comma separated file) is attached to this post.

You assignment is to use R to mine the data for association rules.  Provide information on  all relevant statistics like support, confidence, lift and others.Also, Provide your top 10 rules by lift with the associated metrics. 

```{r}

# Downloading the packages that will be used in this project
library(tidyverse)
library(ggplot2)
library(dplyr)
install.packages("arules", repos = "http://cran.us.r-project.org")
library("arules")
install.packages("arulesViz", repos = "http://cran.us.r-project.org")
library("arulesViz")
```

```{r}
# Getting my working directory
getwd()

# Loading the file in my work space
data<-"/Users/ursulapodosenin/Desktop/GroceryDataSet.csv"
data <- read.transactions(data, sep = ",", format = "basket")
```


```{r}
# Getting an overview of the data
nrow(data)
ncol(data)
summary(data)
```


```{r}
# Calculating the average basket size
basket_sizes <- size(data)
average_basket_size <- mean(basket_sizes)
cat("Average Basket Size:", average_basket_size, "\n")
```


```{r}
# Looking at the distinct number of unique values in the data set
n_items<- unique(unlist(data))
distinct_items<-length(n_items)
number_of_baskets<-nrow(data)
```


```{r}
# Plotting the top 10 most frequently purchased items
itemFrequencyPlot(data, topN = 10, type = "absolute", col = "purple", main = "Top 10 Most Frequently Purchased Items")
```


```{r}
# Creating the association rules
rules <- apriori(data, parameter = list(supp = 0.001, conf = 0.8))
summary(rules)

# Sorting the top 10 rules and saving the rules into a data frame
top_rules <- sort(rules, by = "lift")[1:10]
inspect(top_rules)
top_rules_df <- as(top_rules, "data.frame")
print(top_rules_df)
```

```{r}
# Plotting a graph of the top 10 rules
plot(top_rules, method = "graph", control = list(type = "items"))
```

