---
title: "Weightlifting Analysis"
author: "Ursula Podosenin"
date: "2024-08-10"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

# Loading packages
library(tidyverse)
library(dplyr)
library(ggplot2)
```

```{r}
# Reading the data

getwd()
data<-read.csv("/Users/ursulapodosenin/Desktop/Weightlifitng_2024_Olympics_Data.csv")
head(data)

# Setting Weightclass and Gender as factors
data$Weightclass<-as.factor(data$Weightclass)
data$Gender<-as.factor(data$Gender)
```
### Exploratory Analysis
```{r}

# Plot for Best Snatch excluding missed attempts
ggplot(data[data$Best_Snatch != 0, ], aes(x = Best_Snatch)) +
  geom_histogram(binwidth = 3, fill = "blue", color = "black", alpha = 0.7) +
  facet_wrap(~ Weightclass, scales = "free_x") +
  labs(title = "Distribution of Best Snatch by Weightclass",
       x = "Best Snatch (kg)",
       y = "Count") +
  theme_classic()

# Plot for Best Clean and Jerk excluding missed attempts
ggplot(data[data$Best_CJ != 0, ], aes(x = Best_CJ)) +
  geom_histogram(binwidth = 3, fill = "purple", color = "black", alpha = 0.7) +
  facet_wrap(~ Weightclass, scales = "free_x") +
  labs(title = "Distribution of Clean and Jerk by Weightclass",
       x = "Best Clean (kg)",
       y = "Count") +
  theme_classic()
```
# The distributions of snatches and clean and jerks appear to be irregular with more kurtotic distributions in the snatch and more skewed distributions in the clean and jerk. 
```{r}

# Creating a distribution plot of best snatch by gender
ggplot(data[data$Best_Snatch != 0, ], aes(x = Best_Snatch, fill = factor(Weightclass))) +
  geom_histogram(binwidth = 35, color = "black", position = "dodge") +
  facet_wrap(~ Gender) +
  labs(title = "Distribution of Best Snatch by Weightclass and Gender",
       x = "Best Snatch (kg)",
       y = "Count",
       fill = "Weightclass") +
  theme_classic() +
  scale_fill_brewer(palette = "Set3")

# Creating a distribution plot of best clean and jerk by gender
ggplot(data[data$Best_CJ != 0, ], aes(x = Best_CJ, fill = factor(Weightclass))) +
  geom_histogram(binwidth = 35, color = "black", position = "dodge") +
  facet_wrap(~ Gender) +
  labs(title = "Distribution of Best Clean and Jerk by Weightclass and Gender",
       x = "Best Clean (kg)",
       y = "Count",
       fill = "Weightclass") +
  theme_classic() +
  scale_fill_brewer(palette = "Set3")
```
# The distributions of snatches for the females appear to be normal while the males appear to have a negative skew. For the clean and jerks, the distribution appears to be bimodal for the women and again negatively skewed for the men. 
```{r}

# Finding the total number of people that bombed out in the snatch
num_rows_all_zeroes_snatch<- sum(rowSums(data[, c("First_Attempt_Snatch", "Second_Attempt_Snatch", "Third_Attempt_Snatch")]==0, na.rm=TRUE)== 3&
            !is.na(data$First_Attempt_Snatch) & 
            !is.na(data$Second_Attempt_Snatch) & 
            !is.na(data$Third_Attempt_Snatch))
num_rows_all_zeroes_snatch

num_rows_all_zeroes_snatch
num_rows_all_zeroes_snatch/nrow(data)

# 0.07% of lifters bombed out in the snatch for a total of 9 lifters

# Finding the total number of people that bombed out on the clean and jerk  
num_rows_all_zeros_cj <- sum(rowSums(data[, c("First_Attempt_CJ", "Second_Attempt_CJ", "Third_Attempt_CJ")] == 0, na.rm = TRUE) == 3 & 
                          !is.na(data$First_Attempt_CJ) & 
                          !is.na(data$Second_Attempt_CJ) & 
                          !is.na(data$Third_Attempt_CJ))
num_rows_all_zeros_cj
num_rows_all_zeros_cj/nrow(data)

# 0.05% of lifters bombed out in the clean and jerk for a total of 6 lifters

(num_rows_all_zeroes_snatch+num_rows_all_zeros_cj)/nrow(data)

# 0.1% or of lifters bombed out for a total of 15 out of 122 lifters

```

### Chi-Square Test 
```{r}

# Creating a new column indicating missed second and third attempts for snatch
data <- data %>%
  mutate(Missed_Second_Third_Snatch = ifelse(is.na(Second_Attempt_Snatch) | is.na(Third_Attempt_Snatch), NA,
                                             ifelse(Second_Attempt_Snatch == 0 & Third_Attempt_Snatch == 0, 1, 0)))

# Creating a contingency table excluding rows with NA values
contingency_table <- data %>%
  filter(!is.na(First_Attempt_Snatch) & !is.na(Missed_Second_Third_Snatch)) %>%
  select(First_Attempt_Snatch, Missed_Second_Third_Snatch) %>%
  table()

# Performing the Chi-Square test
chi_square_test <- chisq.test(contingency_table)

# Printing the contingency table and the result of the Chi-Square test
head(contingency_table)
print(chi_square_test)

# Creating a new column indicating missed second and third attempts for clean and jerk
data <- data %>%
  mutate(Missed_Second_Third_CJ = ifelse(is.na(Second_Attempt_CJ) | is.na(Third_Attempt_CJ), NA,
                                         ifelse(Second_Attempt_CJ == 0 & Third_Attempt_CJ == 0, 1, 0)))

# Creating a contingency table excluding rows with NA values
contingency_table_cj <- data %>%
  filter(!is.na(First_Attempt_CJ) & !is.na(Missed_Second_Third_CJ)) %>%
  select(First_Attempt_CJ, Missed_Second_Third_CJ) %>%
  table()

# Performing the Chi-Square test
chi_square_test_cj <- chisq.test(contingency_table_cj)

# Printing the contingency table and the result of the Chi-Square test
head(contingency_table_cj)
print(chi_square_test_cj)
```
# Using a Chi-Square test, from the p-values of p= 0.4159 for the snatch and p= 0.1272 for the clean and jerk, it can safely be said that it is unlikely you will miss your second and third attempts if you miss your first attempt given the significance threshold of p< 0.05.


### Logistic Regression 
```{r}

# Creating new columns indicating if both second and third attempts were missed for snatches and clean and jerks
data$Missed_Second_Third_Snatch <- ifelse(data$Second_Attempt_Snatch == 1 & data$Third_Attempt_Snatch == 1, 1, 0)
data$Missed_Second_Third_CJ <- ifelse(data$Second_Attempt_CJ == 1 & data$Third_Attempt_CJ == 1, 1, 0)

# Removing rows with NA values for snatches and clean and jerks separately
clean_data_snatch <- na.omit(data[, c("First_Attempt_Snatch", "Missed_Second_Third_Snatch")])
clean_data_cj <- na.omit(data[, c("First_Attempt_CJ", "Missed_Second_Third_CJ")])

# Fitting logistic regression models for snatch and clean and jerks
logistic_model_snatch <- glm(Missed_Second_Third_Snatch ~ First_Attempt_Snatch, data = clean_data_snatch, family = binomial)
logistic_model_cj <- glm(Missed_Second_Third_CJ ~ First_Attempt_CJ, data = clean_data_cj, family = binomial)

# Step 4: Calculate the probability of missing both second and third attempts if the first attempt was missed for snatches and clean and jerks
predicted_prob_snatch <- predict(logistic_model_snatch, newdata = data.frame(First_Attempt_Snatch = 1), type = "response")
predicted_prob_cj <- predict(logistic_model_cj, newdata = data.frame(First_Attempt_CJ = 1), type = "response")

# Printing the predicted probabilities
print(paste("Predicted probability for Snatch:", predicted_prob_snatch))
print(paste("Predicted probability for Clean & Jerk:", predicted_prob_cj))

```
# Similarly to the Chi-Square test, the logistic regression shows that there is an extremely small, essentially indicating that the likelihood of missing both the second and third attempts after missing the first attempt is almost zero. The logistic regression model is suggesting that, based on the data, missing the first attempt does not strongly predict missing both the second and third attempts. In other words, if an athlete misses their first attempt, the model estimates that the chances of them missing both subsequent attempts are extremely low.


### Probability 
```{r}

# Filtering out rows with NA values in relevant columns
df_clean <- data |>
  filter(!is.na(First_Attempt_Snatch) & !is.na(Second_Attempt_Snatch))

# Converting attempts to binary (1 if missed, 0 if not)
df_clean <- df_clean |>
  mutate(
    First_Missed = ifelse(First_Attempt_Snatch == 0, 1, 0),
    Second_Missed = ifelse(Second_Attempt_Snatch == 0, 1, 0)
  )

# Filtering to only include rows where the first attempt was missed
df_first_missed <- df_clean |>
  filter(First_Missed == 1)

# Calculating the probability of missing the second attempt given that the first attempt was missed
probability <- mean(df_first_missed$Second_Missed)

# Printing the result
print(probability)

```

# Based on this probability calculation, there is a 39% chance of missing your second and third attempts if you miss your first attempt. 

# To summarize the weightlifting data analysis:

# 1. Histograms of Best Snatch and Best Clean and Jerk:Best Snatch: 
# The distribution of best snatch weights is irregular with a more kurtotic distribution. This indicates variability in the best snatch weights across different weight classes.
# Best Clean and Jerk: The distribution shows a bimodal pattern for women and a negatively skewed distribution for men. This suggests that women’s clean and jerk results have two peaks, while men’s results are skewed to lower weights.

# 2. Bomb-Out Rates
# Snatch: Approximately 0.07% of lifters bombed out (missed all three attempts).
# Clean and Jerk: About 0.1% of lifters bombed out.
# This indicates a relatively low rate of complete failures in both events.

# 3. Chi-Square Tests:
# The Chi-Square tests for both snatch and clean and jerk attempts showed p-values of p= 0.4159 for the snatch and p= 0.1272 for the clean and jerk. This suggests no relationship between missing the first attempt and the likelihood of missing the subsequent attempts.

# 4. Logistic Regression:
# the logistic regression models predicted an extremely low probability of missing both second and third attempts after missing the first attempt. This result contrasts with the Chi-Square tests, indicating that while there is a significant association, the likelihood of failing the subsequent attempts after a first miss is very low according to the model.

# 5. Probability Calculation:
# For lifters who missed their first snatch attempt, the probability of missing the second attempt is approximately 0.39 (or 39%). This probability suggests that there is a moderate chance of missing the second attempt if the first attempt is missed.

### Summary:
# The data shows that while there is a significant association between missing the first attempt and missing subsequent attempts (as per the Chi-Square tests), the logistic regression suggests that this association does not strongly predict the probability of failing the second and third attempts. The calculated probability supports the logistic regression’s conclusion, showing that missing the first attempt does not heavily influence the chances of missing the next attempts.


